`r glue::glue('Initial inferCNV results for {params$library_id}')`

### 2025-04-14 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
WIP: Adding text to the template

Here, we have run `inferCNV` with the following settings:

* With two references, `all-immune` and `subset-immune`
  * The former contains all consensus immune cells, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
  * The latter is a subset of those immune cells considering only macrophage and T-cell cell types.

@@

#   theme(
#     plot.margin = margin(rep(20, 4)),
#     axis.text = element_blank(),
#     axis.ticks = element_blank(),
#     panel.grid = element_blank(),
#     aspect.ratio = 1
#   )
```

@@

Set paths to input and output directories and files in the chunk below.

```{r paths}
# cnv validations
validation_tsv <- file.path(ref_dir, "cnv-validation.tsv")

# references themselves so we know which cells belong to reference
immune_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-all-immune.rds")
subset_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-subset-immune.rds")

@@

# cell types
ewings_tsv <- file.path(
  data_dir,
  "results",
  "cell-type-ewings",
  "SCPCP000015",
  params$sample_id,
  glue::glue("{params$library_id}_ewing-celltype-assignments.tsv")
)

@@

- we also need to read in the cnv-validations.tsv

- section 1: what are the cnvs that they detected? do this from the hmm output at least
- section 2: what is the relationship between consensus cell types and the cnvs? is there one?
- section 3: what are the clusters? how big are they? what is their relationship to the cnv scores? do we want to show them on a UMAP?

@@

```{r message = FALSE}
validation_df <- readr::read_tsv(validation_tsv)

@@

# Separate normal reference from query cells
# cnv_query_df <- infercnv_df |>
#  dplyr::filter(
#    stringr::str_starts(cell_id, params$library_id)
#  ) |>
#  dplyr::mutate(
#    barcodes = stringr::str_remove(cell_id, glue::glue("{params$library_id}-"))
#  )
# cnv_reference_df <- dplyr::anti_join(infercnv_df, cnv_query_df)

# Read SCE and create data frame with full results
sce <- readr::read_rds(sce_file)

cnv_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # add in hmm results
  dplyr::left_join(infercnv_df, by = "barcodes") |>
  # add in cell type results
  dplyr::left_join(celltype_df)

rm(sce)

# remove MT and GL chromosomes
chr_columns_to_keep <- which(
  !stringr::str_detect(
    names(cnv_df),
    "chrMT|chrGL000219.1|chrGL000194.1"
  )
)
cnv_df <- cnv_df[, chr_columns_to_keep]
```

@@

## Distributions of CNVs per cell

First, the total number of chromosomes with CNVs, with the goal of comparing reference to query cells.
We hope for a bimodal distribution where reference cells are near zero, and higher values for putative tumor cells in the query set.

@@

per_cell_cnv_df <- cnv_df |>
  dplyr::mutate(total_cnv_per_cell = rowSums(dplyr::across(has_cnv_cols))) |>
  dplyr::select(
    barcodes,
    UMAP1,
    UMAP2,
    infercnv_cluster = subcluster,
    total_cnv_per_cell,
    consensus_annotation,
    ewing_annotation,
    normal_reference,
    is_immune_reference,
    is_subset_reference
  )
```

@@

wrap_plots(all_plot, subset_plot, guides = "collect") +
  plot_annotation(title = "Total CNVs per cell") &
  theme(legend.position = "bottom")
```
Now, let's look at this total CNV per cell but across cell types, using our two sets:

@@

```{r fig.width = 10, fig.height = 5}
### UMAPs of the cell types as a guide ####
# TODO: colors could be more matched but may not be worth it?

ewings_umap <- per_cell_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(
    title = "Ewings cell type",
    color = "annotation"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

consensus_umap <- per_cell_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(consensus_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(
    title = "Consensus cell type",
    color = "annotation"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

wrap_plots(ewings_umap, consensus_umap)
```

```{r fig.width=10}
ggplot(per_cell_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  facet_wrap(vars(normal_reference)) +
  labs(title = "Total CNV per cell")

cells_per_cluster <- per_cell_cnv_df |>
  dplyr::count(infercnv_cluster, normal_reference)

# How many cells per cluster?
ggplot(cells_per_cluster) +
  aes(
    x = infercnv_cluster, # keep in default order so "reference" and "unknown" are grouped
    y = n
  ) +
  geom_col() +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# How many CNV per cell per cluster?
ggplot(per_cell_cnv_df) +
  aes(x = infercnv_cluster, y = total_cnv_per_cell) +
  geom_point(size = 2) +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

>>>>

This notebook explores `inferCNV` results run on library `params$library_id` from `SCPCP000015`.
`inferCNV` was run with two different normal references:

* `all-immune`: All consensus immune cells across `SCPCP000015` libraries, _excluding_ any which were also identified as tumor in the `cell-type-ewings` analysis module
* `subset-immnune`: A subset of `all-immune`, considering only macrophage and T-cell cell types

`inferCNV` was run with the following settings:

* We used the `i3` HMM which has improved run times relative to their `i6` HMM
* We used the default `analysis_mode = "subclusters"`, which allows the HMM to call CNVs at the level of clusters rather than whole samples or individual cells
* We used the following clustering parameters:
  * Leiden clustering run on PCA (not the expression matrix)
  * `k=20` nearest neighbors with a `resolution` parameter of 1 and using the `modularity` objective function

@@

umap_theme <- list(
  coord_fixed(),
  theme_classic(),
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
)
```

@@

```{r paths}
# TSV with expected Ewing CNVs
validation_tsv <- file.path(ref_dir, "cnv-validation.tsv")

# References themselves so we know which cells belong to reference
immune_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-all-immune.rds")
subset_ref_sce_file <- file.path(ref_dir, "normal-references", "SCPCP000015", "ref-subset-immune.rds")

@@

# ewings and consensus cell types
ewings_tsv <- file.path(
  data_dir,
  "results",
  "cell-type-ewings",
  "SCPCP000015",
  params$sample_id,
  glue::glue("{params$library_id}_ewing-celltype-assignments.tsv")
)

@@

@@

```{r message = FALSE}
validation_df <- readr::read_tsv(validation_tsv) |>
  dplyr::filter(diagnosis == "Ewing sarcoma")

@@

# Read SCE and create data frame with full results
cnv_df <- readr::read_rds(sce_file) |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # add in hmm results
  dplyr::left_join(infercnv_df, by = "barcodes") |>
  # add in cell type results
  dplyr::left_join(celltype_df) |>
  # remove MT and GL chromosomes
  dplyr::select(
    -contains("chrMT"),
    -contains("chrGL")
  )
```

@@

## Distribution of CNVs per cell

This section contains figures showing the total number of CNVs detected per cell.

### CNVs in reference vs query cells

The figures below compare the number of CNVs per cell between reference and query cells. 
Ideally, we'll see a bimodal distribution where reference cells are at/near zero, and there is a peak of higher values which may represent tumor cells.
In the density plots below, distributions are colored based on whether the cells are in the reference or not (aka, query cells).

@@

per_cell_cnv_df <- cnv_df |>
  dplyr::mutate(total_cnv_per_cell = rowSums(dplyr::across(dplyr::all_of(has_cnv_cols)))) |>
  dplyr::rename(infercnv_cluster = subcluster)
```

@@

wrap_plots(all_plot, subset_plot, guides = "collect") +
  plot_annotation(title = "Total CNVs per cell") &
  theme(legend.position = "bottom")
```

### CNVs across cell types

Next, we'll look at the per-cell distributions of CNVs across cell types.
Specifically, we'll look at both the cell types from the `cell-type-consensus` module and the `cell-type-ewings` module separately.
Ideally, we'll see that Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.

@@

Next, we'll visualize the total CNV counts as a UMAP.

First, as a guide, let's look at the top 10 `cell-type-ewings` cell types.

```{r}
per_cell_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(color = "annotation") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```
Next, the total CNVs per cell; ideally, higher values will overlap with regions in the UMAP containing putative tumor cells.

```{r fig.width=10}
ggplot(per_cell_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  facet_wrap(vars(normal_reference)) +
  labs(title = "Total CNV per cell") +
  umap_theme
```

## Clusters

Part of `inferCNV`'s algorithm involves clustering the data, and poor clusters have the potential to lead to inaccurate inferences of CNV. 
To see whether we'd like to further tune the clustering parameters, we'll get a size of some of their properties here:

How many cells are there per cluster?
Clusters labeled `reference_` contain only normal reference cells, and clusters labeled `unknown_` contain the query cells.

```{r}
cells_per_cluster <- per_cell_cnv_df |>
  dplyr::count(infercnv_cluster, normal_reference)

ggplot(cells_per_cluster) +
  aes(
    x = infercnv_cluster, # keep in default order so "reference" and "unknown" are grouped
    y = n
  ) +
  geom_col() +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# How many CNV per cell per cluster?
ggplot(per_cell_cnv_df) +
  aes(x = infercnv_cluster, y = total_cnv_per_cell) +
  geom_point(size = 2) +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

We want to look across chrs now, specifically:

```{r}
validation_df
```

```{r}
cnv_validation_df <- per_cell_cnv_df |>
  dplyr::select(
    barcodes,
    infercnv_cluster,
    total_cnv_per_cell,
    normal_reference,
    is_immune_reference,
    is_subset_reference,
    all_of(
      c(
        glue::glue("proportion_scaled_cnv_chr{1:22}"),
        glue::glue("proportion_scaled_loss_chr{1:22}"),
        glue::glue("proportion_scaled_dupli_chr{1:22}")
      )
    )
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, into = c("cnv_type", "chr")) |>
  dplyr::mutate(chr = stringr::str_remove(chr, "^chr"), chr = as.numeric(chr))
```

```{r}
cnv_validation_df |>
  dplyr::filter(normal_reference == "subset-immune") |>
  ggplot() +
  aes(
    x = proportion,
    color = is_subset_reference
  ) +
  geom_density() +
  facet_wrap(vars(chr))

cnv_validation_df |>
  dplyr::group_by(
    normal_reference,
    cnv_type,
    chr
  ) |>
  dplyr::summarize(median_proportion = median(proportion)) |>
  dplyr::arrange(cnv_type)
```
